Notes on conformal manifolds

### 2025-06-17 Draft_MLFlatDirections.tex
correct typos and add flowchart for ASMC

@@

% ---- TIKZ ----
\usepackage{tikz}
\usetikzlibrary{calc}

@@

In this work, we propose a novel approach, where we study directly the scalar potential of supergravity. As already mentionned, the marginal deformation of CFT's in the large N limit, are in correspondance with the flat directions in the potential, i.e. when there is a continuous set a scalar fields such that $\nabla V = 0$. However, the explicit characterization of these flat directions presents formidable technical challenges. Supergravity scalar potentials, even in truncated models, typically involve dozens of scalar fields with intricate non-linear interactions. The resulting expressions for critical points—where all first derivatives vanish—quickly become too complex for traditional symbolic manipulation, rendering analytical approaches computationally intractable. This complexity paradox is particularly frustrating given that physically meaningful solutions often admit relatively simple parametric descriptions once discovered.

The emergence of machine learning techniques in theoretical physics opens new avenues for addressing such complex problems. Instead of solving the full symbolic system analytically from the outset, one can employ numerical methods to sample the solution space and subsequently apply symbolic regression to extract analytical patterns from the data. This hybrid approach has the potential to bypass the computational bottlenecks inherent in purely symbolic methods, while still aiming to uncover exact analytical expressions.

@@

As an initial simplification, we restrict our analysis to the scalar fields $x_{1}, x_{2}, x_{4}, \tilde{x}_{8}$, and $x{10}$. This subset, selected at random from the original set of thirteen scalar fields, serves to render the problem more tractable at the outset. It is important to emphasize, however, that this subset does not constitute a consistent truncation of the full theory. Consequently, the effective theory defined by these five scalars may be formally ill-defined, as higher-order interactions could, in principle, source the remaining fields.

@@

An additional advantage of this approach is its broader applicability. In scenarios where an explicit analytic expression for the potential and its gradient is inaccessible—yet numerical evaluation remains feasible—our method retains its validity. Hence, the numerical strategy adopted here may prove effective even in cases where an exact analytic expression for the potential is out of reach. \bd{other motivation? Example of this ? }

\section{Numerical analysis}
	\subsection{Gradient descent and local analysis}
		%\paragraph{Gradient descent: implementation, sampling, time needed, loss}

		To identify the flat directions in the potential, we will use numerical tools. The procedure is as follows: first, we sample the underlying manifold, and then we use numerical techniques to extract analytical information from the resulting cloud of points. The first step is thus to sample the manifold.

@@

		We then perform gradient descent on the points using TensorFlow’s automatic differentiation. The loss function is defined as:
		\begin{equation}
			\mathcal{L} = \sum_{i = 1}^{n_{points}} ||\left.\nabla V(\vec{X}^{\,i})\,\right|_{\vec{y}=0}||^2
		\end{equation}
		where $\vec{X}^{\,i} = (x^i_1,\dots,x^i_6,\tilde{x}^i_7,\tilde{x}^i_8,\tilde{x}^i_9,x^i_{10},\dots,x^i_{12},\tilde{x}^i_{13})$ and $\vec{y}^{\,i} = (x^i_3,x^i_5,x^i_6,\tilde{x}^i_7,\tilde{x}^i_9,x^i_{11},x^i_{12},\tilde{x}^i_{11})$ denotes the data points. As already mentionned above, we use the analytic formula for $\nabla V$, but we could have perform the gradient descent by fully using automatic differentiation.

@@

		As a first visualization, we present a triangular plot of the data in Fig. \ref{triangular_plot_1_2_4_8_10}. This figure shows all possible 2D projections of the data, along with the 1D histograms of each coordinate after gradient descent.
		\begin{figure}[h!]
			\centering
			\includegraphics[scale = 0.3]{Figures/triangular_plot_1_2_4_8_10.png} %names need to be changed.
			\caption{Triangular plot showing $2d$ projections and $1d$ histograms of the data after the gradient descent}
			\label{triangular_plot_1_2_4_8_10}
		\end{figure}
		We can make a few comments on those plots. First, we observe non-trivial correlations in the data. Some of these may result from larger basins of attraction, such as in the $x_1/x_2$ plot. However, the structures seen in the $x_4/\tilde{x}_8$ or $x_4/x_{10}$ plots likely reflect genuine features of the manifold, and will show latter that this is indeed the case. We also note that all directions appear to be well populated. 
		%Furthermore, we can notice that a large majority of the points converge to the origin $\vec{0}$. This is not surprising, as this just mean that the round sphere solution has a larger basin of attraction, which sounds reasonable. We do not however attempt to shift the points along flat directions to depopulate the origin or repopulate other regions.
		%\paragraph{Local PCA: graphs of number of patches with given local dimension, and with varying size of local patch, discuss optimal patch size (we want at least few points in each direction)}
		Once the gradient descent has been completed and the flat direction sampled, the next step is to identify the structure of the underlying manifold. Our goal is to eventually obtain an analytical expression, not just a numerical description. Before applying symbolic regression to search for such an expression, we can first perform some exploratory analyses to better understand the data. Specifically, we aim to determine the dimensionality of the manifold and wether it consists of a single connected component or multiple disjoint components (e.g., two intersecting hyperplanes).
		To this end, we apply a local Principal Component Analysis (PCA). For each point, we identify its $k$ nearest neighbours and perform a PCA on that local neighbourhood. This procedure allows us to determine how many principal directions are needed to explain a given proportion $\epsilon$ of the data variance. In other words, it provides an estimate of the local dimensionality around each point.
		We perform this analysis for several values of $k$, namely $k \in \{5, 10, 20, 50, 100\}$, and we fix $\epsilon = 0.99$. The results are presented in Fig. \ref{local_pca}.

@@

		In the realm of symbolic regression, the aim is to uncover interpretable mathematical expressions that best describe a given dataset. In the problem discussed in this paper, we are looking for polynomial expressions such that $p(\vec{x}^i) = 0$, with $i \in \{1, \dots, \texttt{n\_{sample}}\}$. This task involves navigating a vast, discrete, and often rugged search space of possible symbolic models, which poses significant challenges for traditional sampling methods. Markov Chain Monte Carlo (MCMC) techniques, while widely used, can struggle with poor mixing and getting trapped in local optima, especially in high-dimensional or multimodal spaces.

		To address these challenges, we employ an Annealed Importance Sampling (AIS) combined with Sequential Monte Carlo (SMC) methods. AIS constructs a sequence of intermediate distributions that smoothly transition from an initial, tractable distribution (e.g., a prior over symbolic expressions) to the complex target posterior distribution. This annealing process is guided by a temperature-like parameter that gradually emphasizes the data likelihood, allowing for more efficient exploration of the probability landscape.

@@

		Once the Annealing loop is over, we end up with a total of $n_{sample}$ particles, which in principle should be close to annihilate our data, but whose coefficient may need some more fine tuning. To deal with it, we ran a quick exploitation phase, where, for each polynomial, we now only modify its coefficients, with a pertubation $\epsilon \sim \mathcal{N}(0, \sigma_1)$, and keep the new particle, only if the Loss function is now getting smaller. This allow to fine-tune the coefficients.

%		\bd{Results for naive choices of parameters (numbers of iteration and particles, probabilities, $\beta$) and motivate them (we want some exploration and then exploitation, not too long computations): for multiple runs we find multiple polynomials (good polynomial: after exploitation we convert the coefficients to integers and square roots, and recompute loss without regularisation, select with threshold). Quantify it nicely: success rates for each polynomials, and absolute number of success, failing rate. Total time needed, without cluster or fancy computers.}
	%\end{itemize}

	The results of this mlethod are as follow. The code finds 8 different following polynomials: 
	\begin{subequations} \label{eq:pols}
    \begin{align}
      p_{1} &= -\sqrt{2}\,x_{1} + \sqrt{2}\,x_{1}x_8 + x_{2}x_8x_{10} - \sqrt{2}\,x_{2}x_{4}x_8,\\
      p_{2} &= 2\,x_{2} - 2\,x_{2}x_8 + \sqrt{2}\,x_{1}x_{10} + 2\,x_{1}x_{4} - x_{2}x_8x_{10}^{2},\\
      p_{3} &= 2\,x_{2} - 2\,x_{2}x_8 + \sqrt{2}\,x_{1}x_8x_{10} + 2\,x_{1}x_{4}x_8
      		- 2\,x_{2} x_{4}^{2}x_8,\\
      p_{4} &= \sqrt{2}\,x_{2} - \sqrt{2}\,x_{2}x_8 + \sqrt{2}\,x_{1}\,x_{4} + x_{1}x_8x_{10}
      		- x_{2}x_{4}x_8x_{10},\\
      p_{5} &= -2\,x_{1} - 2\,x_{2}x_{4}- 2\,x_{1}x_{4}^{2} + 2\,x_{1}x_8 + \sqrt{2}\,x_{2}x_{10}
      		+ x_{1}x_8x_{10}^{2},\\
      p_{6} &= -2\,x_{2}x_{4} - 2\,x_{1}x_{4}^{2} + 2\,x_{2}x_{4}x_8 + \sqrt{2}\,x_{2}x_{10} 
      		- \sqrt{2}\,x_{2}x_8x_{10} +  x_{1}x_8x_{10}^{2},\\
      p_{7} &= -\sqrt{2}\,x_{1}^{2}x_{4} + \sqrt{2}\,x_{2}^{2}x_{4} + \sqrt{2}\,x_{1}x_{2}x_{4}^{2} - x_{1}^{2}x_{10}
      		- x_{2}^{2}x_{10}, \\
      p_{8} &= -2 + 4 x_{8} - 2 x_8^2 + 2 x_{4}^2 x_8 - x_8^2 x_{10}^2
    \end{align}
  \end{subequations}
  where $x_{8} = e^{\tilde{x}_{8}}$. They are found with different frequencies with some polynomials occuring more often than others. The results are presented in \eqref{table:results}.

@@

>>>>

\usepackage[most]{tcolorbox}

@@

% ---- TIKZ ----
\usepackage{tikz}
\usetikzlibrary{calc}
\usetikzlibrary{shapes.geometric, arrows.meta, positioning}

\tikzstyle{startstop} = [ellipse, minimum width=3.5cm, minimum height=1.cm, text centered, draw=black, fill=gray!20]
\tikzstyle{process} = [rectangle, minimum width=6cm, minimum height=1.2cm, text centered, draw=black, fill=myblue!75]
\tikzstyle{decision} = [rectangle, minimum width=5cm, minimum height=1.2cm, text centered, draw=black, fill=myorange!100, aspect=2]
\tikzstyle{arrow} = [thick,->,>=stealth]

@@

In this work, we propose a novel approach, where we study directly the scalar potential of supergravity. As already mentionned, the marginal deformations of CFT's in the large N limit, are in correspondance with the flat directions in the potential, i.e. when there is a continuous set a scalar fields such that $\nabla V = 0$. However, the explicit characterization of these flat directions presents formidable technical challenges. Supergravity scalar potentials, even in truncated models, typically involve dozens of scalar fields with intricate non-linear interactions. The resulting expressions for critical points—where all first derivatives vanish—quickly become too complex for traditional symbolic manipulation, rendering analytical approaches computationally intractable. This complexity paradox is particularly frustrating given that physically meaningful solutions often admit relatively simple parametric descriptions once discovered.

The emergence of machine learning techniques in theoretical physics opens new avenues for addressing such complex problems. Instead of solving the full symbolic system analytically from the outset, one can employ numerical methods to sample the solution space and subsequently apply symbolic regression techniques to extract analytical patterns from the data. This hybrid approach has the potential to bypass the computational bottlenecks inherent in purely symbolic methods, while still aiming to uncover exact analytical expressions.

@@

As an initial simplification, we restrict our analysis to the scalar fields $x_{1}, x_{2}, x_{4}, \tilde{x}_{8}$, and $x_{10}$. This subset, selected at random from the original set of thirteen scalar fields, serves to render the problem more tractable at the outset. It is important to emphasize, however, that this subset does not constitute a consistent truncation of the full theory. Consequently, the effective theory defined by these five scalars may be formally ill-defined, as higher-order interactions could, in principle, source the remaining fields.

@@

An additional advantage of this approach is its broader applicability. In scenarios where an explicit analytic expression for the potential and its gradient is inaccessible—yet numerical evaluation remains feasible—our method retains its validity. In particular, one could replace the analytic gradient by automatic differenciation methods if one can only evaluate the potential. Hence, the numerical strategy adopted here may prove effective even in cases where an exact analytic expression for the potential is out of reach. \bd{other motivation? Example of this ? }

\section{Numerical analysis}
	\subsection{Gradient descent and local analysis}
		To identify the flat directions in the potential, we will use numerical tools. The procedure is as follows: first, we sample the underlying manifold, and then we use numerical techniques to extract analytical information from the resulting cloud of points. The first step is thus to sample the manifold.
			\subsubsection{Sampling the manifold}
		%\paragraph{Gradient descent: implementation, sampling, time needed, loss}

@@

		We then perform gradient descent on the points using TensorFlow’s automatic differentiation \cite{tensorflow2015-whitepaper}. The loss function is defined as:
		\begin{equation}
			\mathcal{L} = \sum_{i = 1}^{n_{points}} ||\left.\nabla V(\vec{X}^{\,i})\,\right|_{\vec{y}=0}||^2
		\end{equation}
		where $\vec{X}^{\,i} = (x^i_1,\dots,x^i_6,\tilde{x}^i_7,\tilde{x}^i_8,\tilde{x}^i_9,x^i_{10},\dots,x^i_{12},\tilde{x}^i_{13})$ and $\vec{y}^{\,i} = (x^i_3,x^i_5,x^i_6,\tilde{x}^i_7,\tilde{x}^i_9,x^i_{11},x^i_{12},\tilde{x}^i_{11})$ denotes the data points. This way we only keep $(x^i_1,x^i_2,x^i_4,\tilde{x}^i_8,x^i_{10})$ alive after we have taken the derivative. As already mentionned above, we use the analytic formula for $\nabla V$, but we could have perform the gradient descent by fully using automatic differentiation.

@@

		As a first visualization, we present a triangular plot of the data in Fig. \ref{triangular_plot_1_2_4_8_10}. This figure shows all possible 2D projections of the data, along with the 1D histograms of each coordinate after gradient descent.
		\begin{figure}[h!]
			\centering
			\includegraphics[scale = 0.3]{Figures/triangular_plot_1_2_4_8_10.png} %names need to be changed.
			\caption{Triangular plot showing $2d$ projections and $1d$ histograms of the data after the gradient descent}
			\label{triangular_plot_1_2_4_8_10}
		\end{figure}
		We can make a few comments on those plots. First, we observe non-trivial correlations in the data. Some of these may result from larger basins of attraction, such as in the $x_1/x_2$ plot. However, the structures seen in the $x_4/\tilde{x}_8$ or $x_4/x_{10}$ plots likely reflect genuine features of the manifold, and will show latter that this is indeed the case. We also note that all directions appear to be well populated. 
		%Furthermore, we can notice that a large majority of the points converge to the origin $\vec{0}$. This is not surprising, as this just mean that the round sphere solution has a larger basin of attraction, which sounds reasonable. We do not however attempt to shift the points along flat directions to depopulate the origin or repopulate other regions.
		%\paragraph{Local PCA: graphs of number of patches with given local dimension, and with varying size of local patch, discuss optimal patch size (we want at least few points in each direction)}
		\subsubsection{Determining the dimension}
		Once the gradient descent has been completed and the flat direction sampled, the next step is to identify the structure of the underlying manifold. Our goal is to eventually obtain an analytical expression, not just a numerical description. Before applying symbolic regression to search for such an expression, we can first perform some exploratory analyses to better understand the data. Specifically, we aim to determine the dimensionality of the manifold and wether it consists of a single connected component or multiple disjoint components (e.g., two intersecting hyperplanes).
		To this end, we apply a local Principal Component Analysis (PCA). For each point, we identify its $k$ nearest neighbours and perform a PCA on that local neighbourhood. This procedure allows us to determine how many principal directions are needed to explain a given proportion $\epsilon$ of the data variance. In other words, it provides an estimate of the local dimensionality around each point.
		We perform this analysis for several values of $k$, namely $k \in \{5, 10, 20, 50, 100\}$, and we fix $\epsilon = 0.99$. The results are presented in Fig. \ref{local_pca}.

@@

		In the realm of symbolic regression, the aim is to uncover interpretable mathematical expressions that best describe a given dataset. In the problem discussed in this paper, we are looking for polynomial expressions such that $p(\vec{x}^{\,i}) = 0$, with $i \in \{1, \dots, \texttt{n\_{sample}}\}$. This task involves navigating a vast, discrete, and often rugged search space of possible symbolic models, which poses significant challenges for traditional sampling methods. Markov Chain Monte Carlo (MCMC) techniques, while widely used, can struggle with poor mixing and getting trapped in local optima, especially in high-dimensional or multimodal spaces.

		\subsubsection{Annealed Sequential Monte Carlo principle}
		To address these challenges, we employ an Annealed Importance Sampling (AIS) combined with Sequential Monte Carlo (SMC) methods. AIS constructs a sequence of intermediate distributions that smoothly transition from an initial, tractable distribution (e.g., a prior over symbolic expressions) to the complex target posterior distribution. This annealing process is guided by a temperature-like parameter that gradually emphasizes the data likelihood, allowing for more efficient exploration of the probability landscape.

@@

		Once the Annealing loop is over, we end up with a total of $n_{sample}$ particles, which in principle should be close to annihilate our data, but whose coefficient may need some more fine tuning. To deal with it, we ran a quick exploitation phase, where, for each polynomial, we now only modify its coefficients, with a pertubation $\epsilon \sim \mathcal{N}(0, \sigma_1)$, and keep the new particle, only if the Loss function is now getting smaller. This allow to fine-tune the coefficients. 
\hspace{-2cm}
\begin{center}
	\begin{minipage}{18cm}
		\begin{tcolorbox}[colback=myblue!5,
			colframe=myblue,
			title=Algorithm flowchart for ASMC algorithm,
			halign title=flush center,
			sharp corners
			]
%	\scalebox{0.5}{
\begin{tikzpicture}[node distance=1.8cm and .5cm]

	% Nodes
	\node (start) [startstop] {Start};

	\node (init_particles) [process, below of=start] {Initialize sparse polynomials};
	\node (set_weights) [process, below of=init_particles] {Initialize weights and variables};
	\node (compute_beta) [process, below of=set_weights] {Compute or adapt temperature $\beta$};
	\node (weight_update) [process, below of=compute_beta] {Compute log-targets and update weights};

	\node (ess_check) [decision, below of=weight_update] {Is ESS $<$ Threshold?};
	\node (resample) [process, right=4.25cm of ess_check] {Resample particles};

	\node (mutation) [process, below of=ess_check, yshift=-0.5cm] {Apply MH mutation to particles};
	\node (acceptance) [process, below of=mutation] {Compute acceptance and update best};

	\node (end_iter) [decision, below of=acceptance] {Is this the last iteration?};

	\node (exploit) [process, below of=end_iter, yshift=-0.5cm] {Optional local search (exploitation)};
	\node (output) [process, below of=exploit] {Return best coefficients and logs};
	\node (stop) [startstop, below of=output] {Stop};

	% Arrows
	\draw [arrow] (start) -- (init_particles);
	\draw [arrow] (init_particles) -- (set_weights);
	\draw [arrow] (set_weights) -- (compute_beta);
	\draw [arrow] (compute_beta) -- (weight_update);
	\draw [arrow] (weight_update) -- (ess_check);

	\draw [arrow] (ess_check) -- node[above] {Yes} (resample);
	\draw [arrow] (resample) |- (mutation);

	\draw [arrow] (ess_check) -- node[left] {No} (mutation);
	\draw [arrow] (mutation) -- (acceptance);
	\draw [arrow] (acceptance) -- (end_iter);

	\draw [arrow] (end_iter) -- node[left] {} ++(-4,0) |- (compute_beta);
	\node [pos=10, below=430pt, xshift=-100pt] {No};
	\draw [arrow] (end_iter) -- node[right] {Yes} (exploit);

	\draw [arrow] (exploit) -- (output);
	\draw [arrow] (output) -- (stop);
		\end{tikzpicture}
		\end{tcolorbox}
\end{minipage}
\end{center}
%		\bd{Results for naive choices of parameters (numbers of iteration and particles, probabilities, $\beta$) and motivate them (we want some exploration and then exploitation, not too long computations): for multiple runs we find multiple polynomials (good polynomial: after exploitation we convert the coefficients to integers and square roots, and recompute loss without regularisation, select with threshold). Quantify it nicely: success rates for each polynomials, and absolute number of success, failing rate. Total time needed, without cluster or fancy computers.}
	%\end{itemize}

	\subsubsection{Results}

	The results of this method are as follow. The code finds 8 different following polynomials: 
	\begin{subequations} \label{eq:pols}
    \begin{align}
      p_{1} &= -\sqrt{2}\,x_{1} + \sqrt{2}\,x_{1}x_8 + x_{2}x_8x_{10} - \sqrt{2}\,x_{2}x_{4}x_8,\\
      p_{2} &= 2\,x_{2} - 2\,x_{2}x_8 + \sqrt{2}\,x_{1}x_{10} + 2\,x_{1}x_{4} - x_{2}x_8x_{10}^{2},\\
      p_{3} &= 2\,x_{2} - 2\,x_{2}x_8 + \sqrt{2}\,x_{1}x_8x_{10} + 2\,x_{1}x_{4}x_8
      		- 2\,x_{2} x_{4}^{2}x_8,\\
      p_{4} &= \sqrt{2}\,x_{2} - \sqrt{2}\,x_{2}x_8 + \sqrt{2}\,x_{1}\,x_{4} + x_{1}x_8x_{10}
      		- x_{2}x_{4}x_8x_{10},\\
      p_{5} &= -2\,x_{1} - 2\,x_{2}x_{4}- 2\,x_{1}x_{4}^{2} + 2\,x_{1}x_8 + \sqrt{2}\,x_{2}x_{10}
      		+ x_{1}x_8x_{10}^{2},\\
      p_{6} &= -2\,x_{2}x_{4} - 2\,x_{1}x_{4}^{2} + 2\,x_{2}x_{4}x_8 + \sqrt{2}\,x_{2}x_{10} 
      		- \sqrt{2}\,x_{2}x_8x_{10} +  x_{1}x_8x_{10}^{2},\\
      p_{7} &= -\sqrt{2}\,x_{1}^{2}x_{4} + \sqrt{2}\,x_{2}^{2}x_{4} + \sqrt{2}\,x_{1}x_{2}x_{4}^{2} - x_{1}^{2}x_{10}
      		- x_{2}^{2}x_{10}, \\
      p_{8} &= -2 + 4 x_{8} - 2 x_8^2 + 2 x_{4}^2 x_8 - x_8^2 x_{10}^2
    \end{align}
  \end{subequations}
  where $x_{8} = e^{\tilde{x}_{8}}$. They are found with different frequencies with some polynomials occuring more often than others. The results are presented in \eqref{table:results}.

@@

 \subsubsection{Setup of ASMC}
